---
title: 'experiment1: airline data analysis'
date: "2023-09-06"
output:
  pdf_document: default
  html_document:
    df_print: paged
---



```{r}
library(nycflights13)
library(tidyverse)
library(janitor)
library(here)
```

#> ── Attaching core tidyverse packages ───────────────────── tidyverse 2.0.0 ──
#> ✔ dplyr     1.1.3     ✔ readr     2.1.4
#> ✔ forcats   1.0.0     ✔ stringr   1.5.0
#> ✔ ggplot2   3.4.3     ✔ tibble    3.2.1
#> ✔ lubridate 1.9.2     ✔ tidyr     1.3.0
#> ✔ purrr     1.0.2     
#> ── Conflicts ─────────────────────────────────────── tidyverse_conflicts() ──
#> ✖ dplyr::filter() masks stats::filter()
#> ✖ dplyr::lag()    masks stats::lag()
#> ℹ Use the conflicted package (<http://conflicted.r-lib.org/>) to force all conflicts to become errors
#> 

```{r}
nycflights13::flights
df1 = flights
glimpse(df1)
```


```{r}
# task 1: flights from Miami in September
df2= df1 |> filter(month == 9, origin == "MIA")

# task 2: flights in September going to Miami
df3 = df1 |> filter(month == 9, dest == "MIA")

# task 2a: flights in September going to MAI
df3a = df1 |> filter(month == 9, dest == "MAI")

# task 3: flights in January going to Miami
df4 = df1 |> filter(month == 1, dest == "MIA")

# task 4: flights in Summer going to Chicago
df5 = df1 |> filter(between(month, 6, 8), dest %in% c("ORD","MDW") )
```




```{r}
# Sometimes flight numbers are repeated, for flights to Miami in September, find the unique flight number
#df6 = df6 |> filter(month == 9, dest == "MIA")
#flight_numbers <- unique(df3$flight)
numbers = unique(df3$flight)
min(numbers)


# create a data frame df6 composed of flights to Miami with a flight number matching the smallest flight number from the set above

df6 = df1 |> filter(dest == "MIA", flight == 83)


#Create a visualization (using ggplot) to show the DELAYS associated with this flight number

ggplot(data = df6,
       mapping = aes(x=dep_delay, y =arr_delay)) + geom_point()


```

## is there correlation between departure delays with miami opposed to newark?
## which airport has the least amount of departure delays?


```{r}
df1 |>
  filter(dest=="MIA") |>
count(origin,sort=TRUE)
```

### Is flight time affected by delay departure.

```{r}
df7=df1 |>
filter(dest=="MIA",origin=="LGA") |>
  mutate(flt_delta=arr_delay-dep_delay)
```

```{r}
df7 |>
  ggplot(aes(x=dep_delay,y=flt_delta)) +geom_point(alpha=.1)
```
```{r}
df7 |>
  ggplot(aes(x=dep_delay,y=flt_delta)) +geom_point(alpha=.1)+
  geom_hline(aes(yintercept=mean(flt_delta,na.rm=TRUE)))
```
## is departure delay affected by time of year?

```{r}
df7 |>
  ggplot(aes(x=time_hour,y=dep_delay)) +geom_point(alpha=.1)
```


```{r}
df7 |>
  ggplot(aes(x=time_hour,y=dep_delay)) +
  geom_point(alpha=.1)+
  stat_smooth()+
  ylim(-25,120)
```
## does departure delay change across time of day?
```{r}
df7 |>
  ggplot(aes(x=hour+minute/60,y=dep_delay)) +
  geom_point(alpha=.1)+
  stat_smooth()+
  ylim(-25,120)
```

```{r}
df7 |>
  mutate(day_of_week=weekdays(time_hour)) |>
  ggplot(aes(x=hour+minute/60,y=dep_delay,color=day_of_week)) +
  geom_point(alpha=.1)+
  stat_smooth()+
  ylim(-25,120)
```

```{r}
df7 |>
  mutate(day_of_week=weekdays(time_hour)) |>
  ggplot(aes(x=hour+minute/60,y=dep_delay,color=day_of_week)) +
  geom_point(alpha=.1)+
  stat_smooth()+
  ylim(-20,40)+
facet_wrap(~day_of_week)

```








